Towards an Automated Sportscasting System for Chess

### 2011-05-18 paper/paper.tex
Fixed refs.bib

\paragraph{Deep learning.} These techniques attempt to capture complex relationships between input variables in order to transform the input into a different representation which better expresses the relevant structures that make up the input. Specifically, they compose many layers of non-linearities in order to compactly represent complex functions which can model these relationships. They have been used successfully for many different types of tasks \cite{ng-mnist,ng-vision,ng-nlp}.

\paragraph{Automated sportscasting.} Chen and Mooney (\cite{learning-to-sportscast}) present a commentator system that learns language from sportscasts of simulated RoboCup games. Their system, which is trained on ambiguous supervision in the form of textual human commentaries and simulation states of the soccer games, simultaneously tries to establish correspondences between the commentaries and the simulation states. Their work focuses on generating meaningful English commentary from relatively simple topics such as ‘Player 1 kicked the ball to Player 2’, while our work focuses more on identifying complex topics.

\paragraph{Automatic chess commentary.} There exist techniques for automatic annotation of chess games which are heavily dependent on rule-based logic and use of a chess-engine to determine which parts of a specific chess game are salient~\cite{cambridge-chess-annotation}. As a departure from the rule-based approach to automatic chess commentary, Guid et al. (\cite{learning-positional-features}) discuss the use of argument based machine learning, which analyzes reasoning where arguments for and against a certain claim are produced and evaluated (in the context of chess, why a certain move was played over others.) Their system, much like the one presented in this paper, identifies complex positional features of a given chess game.

@@

\subsection{Preparing the Data}
JINGHAO: Write about slicing and tag generation

\section{Features}
\subsection{Sparse auto-encoder}
The sparse auto-encoder is one of many methods in use for unsupervised feature learning~\cite{single-layer-networks-unsupervised-feature-learning}. It takes the form of a feedforward neural network, which consists of an input layer of N nodes, some number of hidden layers, and output layer of M nodes. The value of each layer $\vec{l_i}$ is determined with a weight matrix $W_i$ and a bias vector $b_i$: \[L_i = f(W_i l_{i-1} + b_i),\] where $f(\cdot)$ is a non-linearity like the sigmoid $f(x) = \frac{1}{1 + e^{-z}}$, applied component-wise.

We want to

\subsection{Manually-generated features}
\begin{table*}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Category} & \textbf{Description} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move.
\hline
\end{tabular} 
\caption{List of chess-specific manual features per position.}
\end{table*}

As a first step towards classification, we selected chess-specific manual features from a given position-move pair.  The position-based features involve information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move is a capturing move or a check.  We list all our hand-selected manual features in Table REF.

@@

\begin{table*}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage with the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage to the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opposing player’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, along with explanations}
\label{table:themes}
\end{table*}

@@

\section{Experiments}
\subsection{SVM Based Structured Classifier}

In our experiments, we processed the input data into multimaps from game prefixes to tags. We then segmented the data into three disjoint sets: 90% for training, 5% as the validation set (for tuning our classifiers), and 5% as the test set. The segmenting was done by randomly sampling from the set of all labeled games.

@@

TALK ABOUT AUTOENCODER RESULTS

@@

@@

stack of work:
Segmented:
True positive: 13 (0.001252)
False positive: 2 (0.000193)
True negative: 10213 (0.983438)
False negative: 157 (0.015118)
Precision/Recall: 0.866667, 0.076471
F1 Score: 0.140541
True positive: 12841 (0.001236)
False positive: 154930 (0.014919)
True negative: 10060070 (0.968712)
False negative: 157159 (0.015133)
Precision/Recall: 0.076539, 0.075535
F1 Score: 0.076034

Unsegmented:
True positive: 11 (0.002087)
False positive: 0 (0.000000)
True negative: 5106 (0.968880)
False negative: 153 (0.029032)
Precision/Recall: 1.000000, 0.067073
F1 Score: 0.125714
True positive: 22387 (0.004248)
False positive: 145688 (0.027645)
True negative: 4960312 (0.941236)
False negative: 141613 (0.026872)
Precision/Recall: 0.133196, 0.136506
F1 Score: 0.134831

\begin{table*}
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{True Positive} & \textbf{False Positive} & \textbf{True Negative} & \textbf{False Negative} & \textbf{Precision} & \textbf{Recall} & \textbf{F1} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move.
\end{tabular} 
\caption{List of chess-specific manual features per position.}
\end{table*}

>>>>

\paragraph{Deep learning.} These techniques attempt to capture complex relationships between input variables in order to transform the input into a different representation which better expresses the relevant structures that make up the input. Specifically, they compose many layers of non-linearities in order to compactly represent complex functions which can model these relationships. They have been used successfully for many different types of tasks \cite{Vincent,Lee,Socher}.

\paragraph{Automated sportscasting.} Chen and Mooney (\cite{chen}) present a commentator system that learns language from sportscasts of simulated RoboCup games. Their system, which is trained on ambiguous supervision in the form of textual human commentaries and simulation states of the soccer games, simultaneously tries to establish correspondences between the commentaries and the simulation states. Their work focuses on generating meaningful English commentary from relatively simple topics such as ‘Player 1 kicked the ball to Player 2’, while our work focuses more on identifying complex topics.

\paragraph{Automatic chess commentary.} There exist techniques for automatic annotation of chess games which are heavily dependent on rule-based logic and use of a chess-engine to determine which parts of a specific chess game are salient~\cite{cambridge-chess-annotation}. As a departure from the rule-based approach to automatic chess commentary, Guid et al. (\cite{Guid}) discuss the use of argument based machine learning, which analyzes reasoning where arguments for and against a certain claim are produced and evaluated (in the context of chess, why a certain move was played over others.) Their system, much like the one presented in this paper, identifies complex positional features of a given chess game.

@@

\subsection{Preparing the Data}
After parsing each game, we filtered out the tagged games that had more than 10 plies because it will be very difficult to identify the tags that apply to a complex sequence of moves. Among the games that we kept, we also kept only the first few positions and moves (the ``prefix'' of each game, as described earlier), a parameter that can be adjusted and is one by default. We then generate tags for each game based on the information from the source websites. For example, games in a file that corresponds to a lesson called ``Double Attack'' are labeled doubleAttack. In addition, some of those games may appear in other lessons, and consequently will have multiple labels. An additional step that we take during experimentation is to segment the games further by the color of the player in addition to the tags of the game in order to account for the white-bias in the training data.

\section{Features}
\subsection{Sparse auto-encoder}
The sparse auto-encoder is one of many methods in use for unsupervised feature learning~\cite{Coates}. It takes the form of a feedforward neural network, which consists of an input layer of N nodes, some number of hidden layers, and output layer of M nodes. The value of each layer $\vec{l_i}$ is determined with a weight matrix $W_i$ and a bias vector $b_i$: \[L_i = f(W_i \cdot \vec{l_{i-1}} + b_i),\] where $f(\cdot)$ is a non-linearity like the sigmoid $f(x) = \frac{1}{1 + e^{-z}}$ applied component-wise.

An auto-encoder is such a neural network trained so as to minimize the error (specifically, the sum-squared difference) between the output layer and the input layer, over some set of input examples. If we have fewer nodes in the hidden layers than in the input or output layer, the neural network is forced to learn a compressed representation of the data (represented by the activations, or outputs, of the smaller number of hidden nodes). Instead, we keep a larger number of hidden nodes while imposing a sparsity constraint that the average activation of each one be small (e.g., 0.05). Other work has shown that such a constraint better forces the auto-encoder to learn interesting structures from the data~\cite{bengio}.

A normal auto-encoder only has one layer of hidden nodes, but we can also add more hidden layers to form a stacked auto-encoder. Per~\cite{Bengio07greedylayer-wise}, we use a greedy strategy to train the extra hidden layers in the auto-encoder. We train an auto-encoder with one hidden layer as usual; once this completes, we obtain the output of the hidden layer over all the examples in the input set, and use that as the new set of examples. This allows us to learn higher-level representations of the data~\cite{Vincent}.

Other strategies for unsupervised feature learning include restricted Boltzmann machines (which are undirected), deep belief nets (stacked RBMs, akin to stacked auto-encoders), and denoising autoencoders (which try to reconstruct some input from a corrupted version of it). We only evaluated sparse stacked auto-encoders in this paper.

\subsection{Manually-generated features}
\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Category} & \textbf{Description} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move. \\
\hline
\end{tabular} 
\caption{List of chess-specific manual features per position.}
\end{table*}

As a point of comparison to auto-encoders, we selected chess-specific manual features from a given position-move pair.  The position-based features involve information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move is a capturing move or a check.  We list all our hand-selected manual features in Table REF.

@@

\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Name} & \textbf{Description} \\ \hline
\tt{attackPossible} & The current player has opportunity to attack. \\
\tt{backrankWeakness} & There exists a future checkmate by a rook or queen on the edges of the board. \\
\tt{bishopPair} & There exists a decisive strategic advantage for the player with two bishops. \\
\tt{clearance} & There exists a maneuver to break the pawn structure. \\ 
\tt{closure} & There exists a maneuver to fix the pawn structure. \\
\tt{decoy} & There exists a maneuver to distract a crucial piece. \\
\tt{defensePressure} & There exists a maneuver to break the opposing King’s defense. \\
\tt{deflection} & There exists a maneuver to deflect an attack.  \\ 
\tt{discoveredAttack} & There exists a maneuver to allow for a discovered attack (an attack that was blocked by a friendly piece.) \\
\tt{doubleAttack} & There exists a move that simultaneously attacks at least two crucial squares. \\
\tt{drawCombinationsStalemate} & There exists a combination that allows for decisive material advantage or immediate checkmate. \\
\tt{enclosedKings} & There exists a maneuver to trap the opposing King. \\
\tt{freeingBlockingSquares} & There exists a move to free a crucial square. \\
\tt{goodEndgameChances} & There exists a move to allow for a favorable endgame. \\
\tt{inBetweenMove} & There exists an interrupting move that allows for decisive material advantage of immediate checkmate. \\ 
\tt{interference} & There exists a maneuver to block a crucial piece. \\ 
\tt{interruption} & There exists a maneuver that interrupts the opponent’s tactics. \\
\tt{kingInCenter} & There exists a decisive strategic advantage with the player with the King in the center of the board. \\
\tt{kingToBadSquares} & There exists a combination that forces the opposing King to move to a bad square. \\
\tt{longDiagonals} & There exists a strategic advantage to the player that controls squares on the long diagonal. \\
\tt{matingAttack} & There exists a decisive mating attack. \\
\tt{openingClosingDiagonals} & There exists a maneuver to clear a diagonal. \\
\tt{openingClosingFiles} & There exists a maneuver to clear a file. \\
\tt{overload} & There exists a combination that makes a defending piece defend too many squares at once. \\
\tt{passedPawn} & There exists a combination that allows for a passed pawn (a pawn with no opposing pawns in front of it) \\
\tt{pawnPromotion} & There exists a combination that allows for a pawn promotion. \\
\tt{perpetualCheck} & There exists a combination that forces a perpetual check/draw. \\
\tt{pin} & There exists a combination that allows for a pin. \\
\tt{pullingPiecesToBadSquares} & There exists a combination that forces an opposing piece to move to a bad square. \\
\tt{queenCapture} & There exists a combination to capture the opposing Queen. \\
\tt{quietMates} & There exists a quiet move (one that looks inconsequential on the surface) leading to mate. \\
\tt{removalOfDefence} & There exists a maneuver to remove an opponent’s defender. \\
\tt{unpinning} & There exists a maneuver to unpin a piece. \\
\tt{wrongBishop} & The opposing player’s only bishop is limited in use because of the pawn structure of the game. \\ \hline
\end{tabular}
\caption{List of high-level themes we attempt to automatically identify, along with explanations}
\label{table:themes}
\end{table*}

@@

\section{Experiments}
\subsection{SVM Based Structured Classifier}
In our experiments, we processed the input data into multimaps from game prefixes to tags. We then segmented the data into three disjoint sets: 90% for training, 5% as the validation set (for tuning our classifiers), and 5% as the test set. The segmenting was done by randomly sampling from the set of all labeled games.

@@

When we delved deeper to analyze the sources of errors and true positives, we arrived at a surprising result. We found that deflection and removalOfDefence are much more likely to cause false negatives than other tags (Table \ref{table:negatives}). In addition, we found that all of the true positives are for the quietMates tag.

\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Category} & \textbf{Description} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move. \\
\hline
\end{tabular} 
\caption{Sources of error in the SVM classifier.}
\label{table:negatives}
\end{table*}

Upon closer examination, it was clear that the “checkmate move” feature contributed to that: Any board that is in a checkmate situation is likely to be a quietMate. This interesting result demonstrates the limitations of pursuing a general approach for a highly-complex and specific game like chess.

We sought to improve upon this baseline with an autoencoder. TALK ABOUT AUTOENCODER RESULTS

@@

\bibliographystyle{acl}
\bibliography{paper}

@@
